---
title-meta: "DATA301-DATA471: Data Science in Practice -- Assignment 4"
author-meta: "Leon Cai"
format:
  pdf:
    documentclass: article
    classoption: [11pt, a4paper]
    pdf-engine: lualatex
    geometry: margin=2.5cm
    number-sections: true
    toc: false
    colorlinks: true
    linkcolor: blue
    citecolor: blue
    urlcolor: blue
    fig-pos: "H"
    fig-width: 7
    fig-height: 4
    fig-align: center
    include-in-header:
      text: |
        \usepackage{parskip}
        \usepackage{microtype}
        \usepackage{titlesec}
        \titleformat{\section}{\normalfont\Large\bfseries}{}{0pt}{}
        \titleformat{\subsection}[block]{\normalfont\large\bfseries}{\thesubsection}{1em}{}
        \setlength{\emergencystretch}{3em}
        \usepackage{enumitem}
        \setlist[enumerate]{leftmargin=1.8em, itemsep=0.25em}
        \setlist[itemize]{leftmargin=1.4em, itemsep=0.25em}
    include-before-body:
      text: |
        \title{\textbf{DATA301-DATA471: Data Science in Practice}\\[0.3em]
        \large Assignment 4}
        \author{Leon Cai (300692424)}
        \date{Victoria University of Wellington \\ September 2026}
        \maketitle
        \vspace{-1em}
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(fpp2)
set.seed(471)
```

# Exercise 1

## Order statistics for the quartiles [10%]

**Summary of the two articles.** Nine common definitions of the sample
quantile are compared in @wicklin2017, and they all follow the same
general form given in @hyndman1996 (H&F): $\hat Q(p) = (1-\gamma)x_{(j)} +
\gamma x_{(j+1)}$, where $x_{(j)}$ and $x_{(j+1)}$ are the order statistics
closest to $p$ and $\gamma$ is an interpolation weight, with $j$ and
$\gamma$ both depending on $p$, the sample size $n$, and a parameter $m$
that is different for each definition ($j=\lfloor np+m\rfloor$).

The nine definitions split into two groups. Types 1-3 are discontinuous,
meaning they jump between values instead of changing smoothly, since they
come directly from the empirical CDF, and they always give a value that
already appears in the data (or the average of two adjacent values) --
this group is plotted as a staircase in @wicklin2017. Types 4-9 are continuous and
interpolate between order statistics using a *plotting position* $p_k$,
the probability assigned to order statistic $x_{(k)}$, and these six types
split further by how $p_k$ is derived: types 6, 7 and 8 use a
*distribution-free* rule ($p_k$ does not depend on the population
distribution $F$), while type 9 is built to be almost unbiased only for
QQ-plots against a Normal distribution, so it depends on $F$ instead.

Each definition is also checked in @hyndman1996 against six properties a
good sample quantile should have (continuity in $p$, symmetry in the
tails, uniqueness, and matching the usual sample median at $p=0.5$), and
only type 5 satisfies all six, but it's a compromise that was not built
from an estimation argument. Types 6-9 each satisfy five of the six and
have a clearer theoretical basis, and of these, H&F recommend type 8 as a
general default because its estimate of $Q(p)$ is approximately
median-unbiased for *any* $F$ -- types 6 and 7 don't have this unbiased
property, and type 9 is unbiased only for the Normal case. There's also a
practical point in @wicklin2017 that matters here: the nine definitions
agree closely once $n$ is fairly large (the article uses $n=100$), but
differ more for small samples or when the data have large gaps between
neighbouring values, which includes the $n=5$ cells used in the Monte
Carlo comparison below.

**Choice used in this report.** Type 8 is used throughout (the default for
`lambda_iqr()` above, and everywhere it is called), for two extra reasons
on top of H&F's own recommendation: the Exponential distribution used
here is clearly skewed, unlike the Normal that type 9 targets, and the
smallest sample size in the design, $n=5$, is exactly where @wicklin2017
shows the choice of definition matters most. And this choice isn't only
about this section -- $\hat\lambda_{IQR}$ gets used again in the
bootstrap correction below, and in every replication of the Monte Carlo
experiment further on, so a bad quantile definition here would carry
through to those results too.

**Software default.** R's `quantile()` defaults to type 7, confirmed
below. This comes from S-Plus, whose `quantile()` used type 7
(@hyndman1996, Section 4), and R inherited it. Type 7 is not the type 8
that H&F recommend.

```{r q1-defaults}
quantile(1:10)                              # uses R's default type
identical(quantile(1:10), quantile(1:10, type = 7)) # confirms default = type 7
```

## Bootstrap bias-corrected estimator [20%]

**Derivation.** The bias of $\hat\lambda_{IQR}$ would ideally be found by
simulating many samples from the true Exponential($\lambda$) distribution
and comparing the average of $\hat\lambda_{IQR}$ across those samples to
the true $\lambda$, as covered in the Week 9 lecture, but in practice
there is only one sample $\mathbf{X}$ and the true $\lambda$ is unknown.
Bootstrap gets around this by treating $\mathbf{X}$ itself as a stand-in
for the
population, and $\hat\lambda_{IQR}(\mathbf{X})$ as a stand-in for the
unknown true value. For a fixed sample $\mathbf{X}$ of size $n$, draw $B$ bootstrap
resamples $\mathbf{X}^*_1,\dots,\mathbf{X}^*_B$ (sampling $n$ values from $\mathbf X$
with replacement). The bootstrap estimate of the bias of $\hat\lambda_{IQR}$ is

$$\widehat{\mathrm{bias}} = \frac{1}{B}\sum_{b=1}^B \hat\lambda_{IQR}(\mathbf{X}_b^*) - \hat\lambda_{IQR}(\mathbf{X}).$$

The first-order bias-corrected estimator subtracts this estimated bias:

$$\tilde\lambda(\mathbf{X}) = \hat\lambda_{IQR}(\mathbf{X}) - \widehat{\mathrm{bias}} = 2\hat\lambda_{IQR}(\mathbf{X}) - \frac{1}{B}\sum_{b=1}^B \hat\lambda_{IQR}(\mathbf{X}_b^*).$$

```{r estimators}
lambda_ml <- function(x) 1 / mean(x)

lambda_iqr <- function(x, type = 8) {
  q <- quantile(x, probs = c(0.25, 0.75), type = type)
  log(3) / (q[[2]] - q[[1]])
}

lambda_tilde <- function(x, B = 200, type = 8) {
  n <- length(x)
  boot_est <- replicate(B, lambda_iqr(sample(x, n, replace = TRUE),
                                       type = type))
  # Small n: a bootstrap resample can tie at the quartiles (q3=q1), making
  # lambda_iqr(.) = Inf. Drop these before averaging so one Inf doesn't
  # dominate the correction.
  boot_est <- boot_est[is.finite(boot_est)]
  2 * lambda_iqr(x, type = type) - mean(boot_est)
}
```

```{r estimators-demo}
x <- rexp(30, rate = 2)
c(ML = lambda_ml(x), IQR = lambda_iqr(x), bias_corrected = lambda_tilde(x))
```

## Monte Carlo comparison [20%]

**Design.** The structure follows the course's own Monte Carlo tutorial
[@frery2026mc]: for each situation $(\lambda, n) \in \Lambda \times N$,
$R$ independent replications are generated, each replication gives one
sample $\mathbf{X}$ of size $n$ and one value of each of the three
estimators, and the MSE for that situation is the sample mean of
$(\hat\lambda - \lambda)^2$ over the $R$ replications.

$R=5000$ is used for every situation here, straight from the Week 9
lecture, where the lecturer said 5000 replications per case is enough
for an assignment this size, and that you only need to control $R$ more
tightly in much bigger experiments. It also leaves a
decent margin on precision: the standard error of the Monte Carlo
estimate of MSE is about
$\sqrt{\widehat{\mathrm{Var}}[(\hat\lambda-\lambda)^2]/R}$, and even at
$n=5$, where the estimators are most variable, $R=5000$ keeps this small
next to the MSE values being compared.

```{r mc-experiment}
#| cache: true
#| fig-width: 7
#| fig-height: 4
mc_mse <- function(lambda, n, R, B = 200, type = 8) {
  est <- matrix(NA_real_, nrow = R, ncol = 3,
                dimnames = list(NULL, c("ML", "IQR", "bias_corrected")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, ] <- c(lambda_ml(x), lambda_iqr(x, type = type),
                  lambda_tilde(x, B = B, type = type))
  }
  colMeans((est - lambda)^2)
}

lambda_grid <- c(0.1, 1, 10)
n_grid <- c(5, 30, 100)
R <- 5000

results <- do.call(rbind, lapply(lambda_grid, function(lam) {
  do.call(rbind, lapply(n_grid, function(n) {
    mse <- mc_mse(lam, n, R = R)
    data.frame(lambda = lam, n = n, R = R, t(mse), check.names = FALSE)
  }))
}))
results
```

The same numbers are easier to compare as a plot, especially since
`bias_corrected`'s MSE spans several orders of magnitude across the
design -- @fig-mc puts both axes on a log scale.

```{r fig-mc}
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "MSE of the three estimators across the $(\\lambda, n)$ design, log-log scale."
est_names <- c("ML", "IQR", "bias_corrected")
results_long <- do.call(rbind, lapply(est_names, function(est) {
  data.frame(lambda = results$lambda, n = results$n,
             estimator = est, mse = results[[est]])
}))

ggplot(results_long, aes(n, mse, color = estimator)) +
  geom_line() + geom_point() +
  scale_x_log10(breaks = c(5, 30, 100)) +
  scale_y_log10() +
  facet_wrap(~ lambda, labeller = label_both) +
  labs(x = "n (log scale)", y = "MSE (log scale)", color = "Estimator")
```

**Discussion.** $\hat\lambda_{ML}$ has the lowest MSE in every one of the
nine cells. This matches theory. It is the maximum likelihood estimator,
so it uses the full sample through the likelihood, and it has the
smallest asymptotic variance among regular estimators. $\hat\lambda_{IQR}$
only uses two order statistics, so it should do worse. The MSE gap between
the two stays in a similar range across the design: $\hat\lambda_{IQR}$'s
MSE is roughly 1.6 to 2.5 times $\hat\lambda_{ML}$'s, at every
$(\lambda, n)$ combination, with no clear trend as $n$ or $\lambda$
change. This roughly matches the idea that the IQR estimator has a fixed
*asymptotic* efficiency loss from throwing away most of the sample, but
that result only has to hold as $n\to\infty$ -- some of the spread in the
ratio seen here could be genuine finite-sample behaviour rather than just
Monte Carlo noise from a design that is too small to pin the ratio down
exactly.

The bias-corrected estimator $\tilde\lambda$ is *worse* than the
uncorrected $\hat\lambda_{IQR}$ at every single $(\lambda, n)$ in the
design, not only at $n=5$. At $n=100$, the correction makes MSE about
15-18% bigger than $\hat\lambda_{IQR}$, across all three $\lambda$
values. At $n=30$, the gap is about 24-26%. At $n=5$, it is much worse:
three to four orders of magnitude bigger. The reason is a bias-variance
trade-off that goes the wrong way. $\hat\lambda_{IQR}$ is a nonlinear
function (a reciprocal) of the sample's interquartile spread. A bootstrap
resample can, by chance, have a smaller or larger spread than the
original sample. After taking the reciprocal, and then doubling
it in $\tilde\lambda = 2\hat\lambda_{IQR} - \overline{\hat\lambda_{IQR}^*}$,
this small chance difference turns into a large swing in the correction
term. The variance this adds is bigger than the bias it removes, so MSE
gets worse overall, and more so as $n$ shrinks, because bootstrap
resamples are then more likely to have an unusually small spread. So the
first-order bootstrap correction does not help this estimator, at any of
the sample sizes tested here.

All three estimators' MSE decreases roughly with $1/n$, matching the
usual parametric convergence rate, and increases roughly with
$\lambda^2$: at $n=5$, $\mathrm{MSE}(\hat\lambda_{ML})/\lambda^2$ is
about 0.64, 0.57, and 0.62 at $\lambda=0.1,1,10$, close enough that the
ratio looks roughly constant. That makes sense: $\hat\lambda$ is a
rate, so squared error scales with $\lambda^2$ under a change of unit.

\newpage

# Exercise 2

`ausbeer` comes with the `fpp2` package. It is quarterly beer production
data for Australia. @fig-ausbeer shows the full series; the sections on
frequency, trend, seasonality, variance and unusual observations all
refer back to this same series, some directly from the plot and some
from numbers computed later.

```{r fig-ausbeer}
#| out-width: "85%"
#| fig-cap: "Quarterly Australian beer production, 1956 Q1--2010 Q2."
autoplot(ausbeer, color = "steelblue") +
  labs(y = "Megalitres")
```

## What is the frequency of the series? How can you determine this from the data and/or the time-series plot?

The frequency is 4 (quarterly data), and the series runs from 1956 Q1 to
2010 Q2, confirmed below by `frequency()`, `start()` and `end()`.

```{r ausbeer-freq}
frequency(ausbeer)
start(ausbeer)
end(ausbeer)
```

Counting cycles by eye on the full 54-year series in @fig-ausbeer is
hard, since each year only takes up a few pixels, but zooming into a
handful of years in @fig-ausbeer-zoom makes the four-per-year pattern
countable directly:

```{r fig-ausbeer-zoom}
#| out-width: "85%"
#| fig-cap: "`ausbeer` zoomed into 1985--1990, showing four quarters per year."
autoplot(window(ausbeer, start = 1985, end = 1990), color = "steelblue") +
  labs(y = "Megalitres")
```

## Describe the overall behaviour of the series. Is there evidence of a long-term trend?

There isn't one single trend here, more like three phases. Production
rises steadily from 1956 to around the mid-1970s (from about 260 to
about 485 megalitres, nearly doubling), then plateaus and fluctuates
around a peak from the mid-1970s to about 1990, then drops fairly
sharply in the early 1990s (from about 490 down to about 445 in just
three years) before levelling off at that lower level through the
2000s.

To check this against a straight line without fitting any model, @fig-ausbeer-trend
joins the early-period average (1956-60) to the late-period average
(2005-10) with a red line. If the trend were linear, the yearly average
would stay close to this line all the way through. Instead it sits above
it for about thirty-five years, from the mid-1960s to around 2000, by
roughly 100-165 megalitres between 1970 and 1990, and only comes back to
the line at the end. The yearly average, in blue, makes the three phases
clearer than either the raw quarterly plot (@fig-ausbeer) or the
straight line on their own.

```{r fig-ausbeer-trend}
#| out-width: "85%"
#| fig-cap: "`ausbeer`'s yearly average (blue) against a straight line joining the early- and late-period averages (red)."
ausbeer_df <- data.frame(time = as.numeric(time(ausbeer)),
                          value = as.numeric(ausbeer))
yearly <- aggregate(value ~ floor(time), ausbeer_df, mean)
names(yearly) <- c("year", "value")
yearly <- yearly[yearly$year < 2010, ]  # 2010 only has 2 quarters, drop it

early <- mean(ausbeer_df$value[ausbeer_df$time < 1961])
late <- mean(ausbeer_df$value[ausbeer_df$time >= 2005])
t_early <- mean(ausbeer_df$time[ausbeer_df$time < 1961])
t_late <- mean(ausbeer_df$time[ausbeer_df$time >= 2005])
ausbeer_df$line <- early +
  (late - early) * (ausbeer_df$time - t_early) / (t_late - t_early)

ggplot(ausbeer_df, aes(time, value)) +
  geom_line(alpha = 0.5) +
  geom_line(aes(y = line), color = "red", linetype = "dashed") +
  geom_line(data = yearly, aes(year + 0.5, value),
            color = "blue", linewidth = 1) +
  labs(x = "Time", y = "Megalitres")
```

## Is there evidence of seasonality? If so, describe the seasonal pattern.

Yes, and it's a strong, regular pattern: every year has the same shape,
low in Q2 and Q3 and high in Q4, confirmed by the quarter averages
below. Q4 is clearly the highest quarter and Q2 the lowest.

```{r ausbeer-quarter-means}
quarter_means <- aggregate(as.numeric(ausbeer) ~ cycle(ausbeer), FUN = mean)
names(quarter_means) <- c("quarter", "mean")
quarter_means
```

@fig-ausbeer-box shows this isn't just an average effect: Q4's box
sits above Q2's and Q3's with no overlap at all, and only a small
overlap with Q1's, so Q4 being the high quarter is a consistent,
year-on-year pattern rather than something driven by a few big years.

```{r fig-ausbeer-box}
#| out-width: "85%"
#| fig-cap: "Distribution of `ausbeer` by quarter."
qdf <- data.frame(quarter = factor(cycle(ausbeer)),
                   value = as.numeric(ausbeer))
ggplot(qdf, aes(quarter, value, fill = quarter)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.08, alpha = 0.3, size = 0.6, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Quarter", y = "Megalitres")
```

## Does the variability appear constant over time? Explain what you see in the plot and whether a transformation might be appropriate.

Not really constant, but not in the usual direction. This shows up in
@fig-ausbeer: the seasonal swings look visibly tighter in the 2000s
than they did in the 1970s-90s, even though the level in that period
isn't dramatically different. The breakdown below puts a number on it. The
variability (measured here as the standard deviation within each
period) actually *shrinks* over time, even though the level itself goes
up and then back down:

```{r ausbeer-variance}
breaks <- c(1955, 1970, 1980, 1990, 2000, 2011)
labels <- c("1956-70", "1971-80", "1981-90", "1991-2000", "2001-10")
period <- cut(ausbeer_df$time, breaks = breaks, labels = labels)
aggregate(value ~ period, data.frame(ausbeer_df, period),
          function(x) c(mean = mean(x), sd = sd(x)))
```

The standard deviation doesn't track the mean in any simple way here:
1971-80 has the highest sd despite a middling mean, and 1981-90 has the
highest mean despite one of the *lower* sds. The clearest pattern is at
the far end, where the most recent period's spread (38) is distinctly
smaller than every earlier period's (54-61), even though its own level
isn't unusual. The relative variability (sd divided by mean) tells a
similar story, highest early and lowest late, though not a clean
monotonic slide in between. Either way, this is not the classic case
for a log transform, where the size of the fluctuations is supposed to
grow *with* the level. If anything the opposite shows up at the two
ends of the series.

## Are there any observations that you would regard as unusual? Identify them and explain why they attract attention.

Two things stand out, of different kinds. The first is structural: the
whole pre-1975-ish period behaves differently from the rest of the
series, both in level (much lower) and in relative variability (much
higher, as shown above). Even the "outlier" points below the whiskers
in @fig-ausbeer-box are all from 1956-61, so they're this same
structural thing showing up again rather than scattered noise. This
same split shows up in @hyndman2018, which restricts to
`window(ausbeer, start = 1992)` before fitting any forecasting model to
this dataset, since the earlier decades reflect a different, still
growing market rather than the more settled pattern seen from the 1990s
on.

The second is a genuine point-level anomaly: 1981 Q1 comes in at 548,
well above its neighbouring Q1 values (453-515 across the surrounding
decade). Comparing every quarter with the average of the same quarter in the
three years either side (no model involved), it is the largest gap
anywhere in `ausbeer`: 56 megalitres above its neighbours, about 3.8
times the typical size of that gap:

```{r ausbeer-local-gap}
ausbeer_df$quarter <- as.numeric(cycle(ausbeer))
local_gap <- sapply(seq_len(nrow(ausbeer_df)), function(i) {
  near <- ausbeer_df$quarter == ausbeer_df$quarter[i] &
    abs(ausbeer_df$time - ausbeer_df$time[i]) <= 3 &
    seq_len(nrow(ausbeer_df)) != i
  if (sum(near) < 4) return(NA)
  ausbeer_df$value[i] - mean(ausbeer_df$value[near])
})
ausbeer_df$time[which.max(abs(local_gap))]
max(abs(local_gap), na.rm = TRUE) / sd(local_gap, na.rm = TRUE)
```

## Plot the series by season (for example, using a seasonal plot). What does this reveal that is less obvious in the ordinary time-series plot?

```{r fig-ausbeer-season}
#| fig-height: 5
#| out-width: "85%"
#| fig-format: png
#| fig-cap: "Seasonal plot of `ausbeer`, one line per year."
ggseasonplot(ausbeer)
```

@fig-ausbeer-season puts every year on its own line across the four
quarters. Compared to the ordinary time plot (@fig-ausbeer), this makes
a few things easier to see:

- Almost every year has the same up-down-up shape: climbing to a peak
  at Q4 without exception, and usually dipping lowest at Q2 (though Q3
  is the lowest quarter in six of the 54 full years, like 1978 and
  1990).
- The lines are stacked in the same order as the trend. The earliest
  (coolest-coloured) years sit lowest, the lines climb up through the
  1970s-80s, and the most recent years sit back down again.
- Comparing the same quarter across different years is much easier
  here than in the ordinary time plot, where the years are stretched
  out left to right instead of stacked on top of each other.
- The 1981 anomaly from section 2.5 is visible here too: its line
  starts at Q1 clearly above every neighbouring year's line, standing
  out precisely because the rest of that colour range clusters so
  tightly together.

## Plot the seasonal subseries. Which quarters have the highest and lowest typical values? Is the seasonal pattern stable over time?

```{r fig-ausbeer-subseries}
#| out-width: "85%"
#| fig-format: png
#| fig-cap: "Seasonal subseries plot of `ausbeer`, one panel per quarter."
p_subseries <- ggsubseriesplot(ausbeer)
p_subseries$layers[[1]]$aes_params$colour <- "steelblue"
p_subseries$layers[[2]]$aes_params$colour <- "red"
p_subseries + labs(y = "Megalitres")
```

@fig-ausbeer-subseries splits the series into one panel per quarter,
keeping the years in time order within each panel (unlike the jittered
points in @fig-ausbeer-box) and marking each quarter's mean in red:

- Unlike the boxplot, keeping the time order visible means each panel
  still traces its own version of the rise-then-plateau-then-decline
  shape from the trend. So the *level* within a quarter is not stable
  over time, since the seasonal and trend components aren't independent
  of when in the series you're looking.
- It also confirms two things from section 2.3: Q4 has the highest
  typical value and Q2 the lowest, and the seasonal *pattern* (which
  quarter is high and which is low) looks stable, not flipping or
  fading out in any panel.

## Would you detrend or transform the series before modelling it? Give a visual argument for your answer. Do not fit a model.

Detrending, yes. @fig-ausbeer-resid plots the gap between the raw
series and the same straight line as in @fig-ausbeer-trend: just two
averages connected by a line, not a fitted model. If a straight line
described the trend, the gap would only wobble around zero with the
seasonal swings. Instead it forms a clear hump: close to zero at the two
ends, where the line is anchored, then well above zero through the
1970s-80s (on a yearly average, roughly 100-165 megalitres) before
coming back down. That's the same rise-plateau-fall shape as the trend
itself, so a straight line clearly doesn't describe the series, and the
trend would need to come out before it's modelled as something
stationary.

```{r fig-ausbeer-resid}
#| out-width: "85%"
#| fig-cap: "Gap between `ausbeer` and the straight line in Figure 4."
ausbeer_df$gap <- ausbeer_df$value - ausbeer_df$line

ggplot(ausbeer_df, aes(time, gap)) +
  geom_line(color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Time", y = "Value minus the straight line")
```

A variance-stabilising transform (log or Box-Cox) is less clearly
justified. The usual visual argument for a log transform is that spread
grows with the level, and neither @fig-ausbeer nor the breakdown in
section 2.4 really show that. The spread bounces around in the 53-61
range while the mean rises, peaks, and starts to fall, and only drops
noticeably (to 38) in the very last period. So a log transform doesn't
have an obvious visual justification here, even though detrending
clearly does.

\newpage

# References

::: {#refs}
:::

\newpage

# Appendix: Declaration of AI Use {.unnumbered}

Claude Code was used to help check and improve the R code for both
exercises, to summarise and search reference material on Monte Carlo
method applications, to act as a sounding board for discussing ideas,
and to assist with Quarto/\LaTeX{} typesetting and English language
polishing. All written interpretations and analytical decisions are my
own work.
