---
title-meta: "DATA301-DATA471: Data Science in Practice -- Assignment 4"
author-meta: "Leon Cai"
format:
  pdf:
    documentclass: article
    classoption: [11pt, a4paper]
    pdf-engine: lualatex
    geometry: margin=2.5cm
    number-sections: true
    toc: false
    colorlinks: true
    linkcolor: blue
    citecolor: blue
    urlcolor: blue
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{parskip}
        \usepackage{microtype}
        \setlength{\emergencystretch}{3em}
        \usepackage{enumitem}
        \setlist[enumerate]{leftmargin=1.8em, itemsep=0.25em}
        \setlist[itemize]{leftmargin=1.4em, itemsep=0.25em}
    include-before-body:
      text: |
        \title{\textbf{DATA301-DATA471: Data Science in Practice}\\[0.3em]
        \large Assignment 4}
        \author{Leon Cai (300692424)}
        \date{Victoria University of Wellington \\ September 2026}
        \maketitle
        \vspace{-1em}
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(fpp2)
set.seed(471)
```

# Exercise 1

We have i.i.d. samples $\mathbf{X}=(X_1,\dots,X_n)$ from an Exponential distribution
with rate $\lambda>0$ (so $\lambda$ is the reciprocal of the mean). We compare three
estimators of $\lambda$: the maximum likelihood estimator $\hat\lambda_{ML}$, the
inter-quartile-range estimator $\hat\lambda_{IQR}$, and a bootstrap bias-corrected
version of the latter, $\tilde\lambda$.

**Note.** The assignment PDF prints $\hat\lambda_{ML}(\mathbf{X}) = \tfrac1n\sum_i X_i$, which is
just the sample mean $\bar X$ (an estimator of $1/\lambda$, not of $\lambda$). Since the
question statement itself defines $\lambda$ as "the reciprocal of the mean," and the
true MLE of the rate of an Exponential distribution is $\hat\lambda_{ML}=1/\bar X$, we
treat this as a typo in the brief and use $\hat\lambda_{ML}(\mathbf{X}) = 1/\bar X =
n / \sum_i X_i$ throughout.

## Order statistics for the quartiles [10%]

TODO: summary of @wicklin2017 and @hyndman1996 — discrete (type 1-3) vs
interpolation-based (type 4-9) definitions, H&F's recommendation of type 8, and the
default types used by R (`quantile()`, type 7) and Python (`numpy.percentile`,
`method="linear"`, equivalent to type 7).

```{r q1-defaults}
quantile(1:10)                              # uses R's default type
identical(quantile(1:10), quantile(1:10, type = 7)) # confirms default = type 7
```

## Bootstrap bias-corrected estimator [20%]

**Derivation.** For a fixed sample $\mathbf{X}$ of size $n$, draw $B$ bootstrap
resamples $\mathbf{X}^*_1,\dots,\mathbf{X}^*_B$ (sampling $n$ values from $\mathbf X$
with replacement). The bootstrap estimate of the bias of $\hat\lambda_{IQR}$ is

$$\widehat{\mathrm{bias}} = \frac{1}{B}\sum_{b=1}^B \hat\lambda_{IQR}(\mathbf{X}_b^*) - \hat\lambda_{IQR}(\mathbf{X}).$$

The first-order bias-corrected estimator subtracts this estimated bias:

$$\tilde\lambda(\mathbf{X}) = \hat\lambda_{IQR}(\mathbf{X}) - \widehat{\mathrm{bias}} = 2\hat\lambda_{IQR}(\mathbf{X}) - \frac{1}{B}\sum_{b=1}^B \hat\lambda_{IQR}(\mathbf{X}_b^*).$$

```{r estimators}
lambda_ml <- function(x) 1 / mean(x)

lambda_iqr <- function(x, type = 8) {
  q <- quantile(x, probs = c(0.25, 0.75), type = type)
  log(3) / (q[[2]] - q[[1]])
}

lambda_tilde <- function(x, B = 200, type = 8) {
  n <- length(x)
  boot_est <- replicate(B, lambda_iqr(sample(x, n, replace = TRUE), type = type))
  # For very small n, a bootstrap resample can tie at the quartiles (q3 = q1),
  # making lambda_iqr(.) = Inf; we drop these degenerate resamples from the
  # bootstrap average rather than let a single Inf dominate the correction.
  boot_est <- boot_est[is.finite(boot_est)]
  2 * lambda_iqr(x, type = type) - mean(boot_est)
}
```

```{r estimators-demo}
x <- rexp(30, rate = 2)
c(ML = lambda_ml(x), IQR = lambda_iqr(x), bias_corrected = lambda_tilde(x))
```

## Monte Carlo comparison [20%]

TODO: justify the number of Monte Carlo replications per $(\lambda, n)$ cell using
the CLT-based precision argument in @robert2009 (Ch. 3): the standard error of a
Monte Carlo estimate of $\mathrm{MSE}$ is $\sqrt{\widehat{\mathrm{Var}}/R}$, so $R$
is chosen so this is small relative to the MSE values being compared. Smaller $n$
and more extreme $\lambda$ give higher-variance estimators and need larger $R$.

```{r mc-experiment}
#| cache: true
mc_mse <- function(lambda, n, R, B = 200, type = 8) {
  est <- matrix(NA_real_, nrow = R, ncol = 3,
                dimnames = list(NULL, c("ML", "IQR", "bias_corrected")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, ] <- c(lambda_ml(x), lambda_iqr(x, type = type), lambda_tilde(x, B = B, type = type))
  }
  colMeans((est - lambda)^2)
}

lambda_grid <- c(0.1, 1, 10)
n_grid <- c(5, 30, 100)
R_grid <- c(`5` = 3000, `30` = 1500, `100` = 800)

results <- do.call(rbind, lapply(lambda_grid, function(lam) {
  do.call(rbind, lapply(n_grid, function(n) {
    mse <- mc_mse(lam, n, R = R_grid[[as.character(n)]])
    data.frame(lambda = lam, n = n, R = R_grid[[as.character(n)]],
               t(mse), check.names = FALSE)
  }))
}))
results
```

TODO: discuss the table above (which estimator wins in each corner of the design,
whether the bias correction actually helps $\hat\lambda_{IQR}$, small-$n$ vs large-$n$
behaviour).

# Exercise 2

```{r ausbeer}
autoplot(ausbeer) + labs(title = "Quarterly Australian beer production", y = "Megalitres")
```

## Frequency

TODO

## Trend

TODO

## Seasonality

TODO

## Variance stability

TODO

## Unusual observations

TODO

```{r ausbeer-season}
ggseasonplot(ausbeer)
```

## Seasonal plot

TODO

```{r ausbeer-subseries}
ggsubseriesplot(ausbeer)
```

## Seasonal subseries plot

TODO

## Detrend / transform?

TODO

# References
