---
title-meta: "DATA301-DATA471: Data Science in Practice -- Assignment 4"
author-meta: "Leon Cai"
format:
  pdf:
    documentclass: article
    classoption: [11pt, a4paper]
    pdf-engine: lualatex
    geometry: margin=2.5cm
    number-sections: true
    toc: false
    colorlinks: true
    linkcolor: blue
    citecolor: blue
    urlcolor: blue
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{parskip}
        \usepackage{microtype}
        \setlength{\emergencystretch}{3em}
        \usepackage{enumitem}
        \setlist[enumerate]{leftmargin=1.8em, itemsep=0.25em}
        \setlist[itemize]{leftmargin=1.4em, itemsep=0.25em}
    include-before-body:
      text: |
        \title{\textbf{DATA301-DATA471: Data Science in Practice}\\[0.3em]
        \large Assignment 4}
        \author{Leon Cai (300692424)}
        \date{Victoria University of Wellington \\ September 2026}
        \maketitle
        \vspace{-1em}
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(fpp2)
set.seed(471)
```

# Exercise 1

## Order statistics for the quartiles [10%]

**Summary of the two articles.** Nine common definitions of the sample
quantile are compared in @wicklin2017, and they all follow the same
general form given in @hyndman1996 (H&F): $\hat Q(p) = (1-\gamma)x_{(j)} +
\gamma x_{(j+1)}$, where $x_{(j)}$ and $x_{(j+1)}$ are the order statistics
closest to $p$ and $\gamma$ is an interpolation weight, with $j$ and
$\gamma$ both depending on $p$, the sample size $n$, and a parameter $m$
that is different for each definition ($j=\lfloor np+m\rfloor$).

The nine definitions split into two groups. Types 1-3 are discontinuous,
meaning they jump between values instead of changing smoothly, since they
come directly from the empirical CDF, and they always give a value that
already appears in the data (or the average of two adjacent values) --
this group is plotted as a staircase in @wicklin2017. Types 4-9 are continuous and
interpolate between order statistics using a *plotting position* $p_k$,
the probability assigned to order statistic $x_{(k)}$, and these six types
split further by how $p_k$ is derived: types 6, 7 and 8 use a
*distribution-free* rule ($p_k$ does not depend on the population
distribution $F$), while type 9 is built to be almost unbiased only for
QQ-plots against a Normal distribution, so it depends on $F$ instead.

Each definition is also checked in @hyndman1996 against six properties a
good sample quantile should have (continuity in $p$, symmetry in the
tails, uniqueness, and matching the usual sample median at $p=0.5$), and
only type 5 satisfies all six, but it's a compromise that was not built
from an estimation argument. Types 6-9 each satisfy five of the six and
have a clearer theoretical basis, and of these, H&F recommend type 8 as a
general default because its estimate of $Q(p)$ is approximately
median-unbiased for *any* $F$ -- types 6 and 7 don't have this unbiased
property, and type 9 is unbiased only for the Normal case. There's also a
practical point in @wicklin2017 that matters here: the nine definitions
agree closely once $n$ is fairly large (the article uses $n=100$), but
differ more for small samples or when the data have large gaps between
neighbouring values, which includes the $n=5$ cells used in the Monte
Carlo comparison below.

**Choice used in this report.** Type 8 is used throughout (the default for
`lambda_iqr()` above, and everywhere it is called), for two extra reasons
on top of H&F's own recommendation: the Exponential distribution used
here is clearly skewed, unlike the Normal that type 9 targets, and the
smallest sample size in the design, $n=5$, is exactly where @wicklin2017
shows the choice of definition matters most. And this choice isn't only
about this section -- $\hat\lambda_{IQR}$ gets used again in the
bootstrap correction below, and in every replication of the Monte Carlo
experiment further on, so a bad quantile definition here would carry
through to those results too.

**Software default.** R's `quantile()` defaults to type 7, confirmed
below. This comes from S-Plus, whose `quantile()` used type 7
(@hyndman1996, Section 4), and R inherited it. Type 7 is not the type 8
that H&F recommend.

```{r q1-defaults}
quantile(1:10)                              # uses R's default type
identical(quantile(1:10), quantile(1:10, type = 7)) # confirms default = type 7
```

## Bootstrap bias-corrected estimator [20%]

**Derivation.** The bias of $\hat\lambda_{IQR}$ would ideally be found by
simulating many samples from the true Exponential($\lambda$) distribution
and comparing the average of $\hat\lambda_{IQR}$ across those samples to
the true $\lambda$, as covered in the Week 9 lecture, but in practice
there is only one sample $\mathbf{X}$ and the true $\lambda$ is unknown.
Bootstrap gets around this by treating $\mathbf{X}$ itself as a stand-in
for the
population, and $\hat\lambda_{IQR}(\mathbf{X})$ as a stand-in for the
unknown true value. For a fixed sample $\mathbf{X}$ of size $n$, draw $B$ bootstrap
resamples $\mathbf{X}^*_1,\dots,\mathbf{X}^*_B$ (sampling $n$ values from $\mathbf X$
with replacement). The bootstrap estimate of the bias of $\hat\lambda_{IQR}$ is

$$\widehat{\mathrm{bias}} = \frac{1}{B}\sum_{b=1}^B \hat\lambda_{IQR}(\mathbf{X}_b^*) - \hat\lambda_{IQR}(\mathbf{X}).$$

The first-order bias-corrected estimator subtracts this estimated bias:

$$\tilde\lambda(\mathbf{X}) = \hat\lambda_{IQR}(\mathbf{X}) - \widehat{\mathrm{bias}} = 2\hat\lambda_{IQR}(\mathbf{X}) - \frac{1}{B}\sum_{b=1}^B \hat\lambda_{IQR}(\mathbf{X}_b^*).$$

```{r estimators}
lambda_ml <- function(x) 1 / mean(x)

lambda_iqr <- function(x, type = 8) {
  q <- quantile(x, probs = c(0.25, 0.75), type = type)
  log(3) / (q[[2]] - q[[1]])
}

lambda_tilde <- function(x, B = 200, type = 8) {
  n <- length(x)
  boot_est <- replicate(B, lambda_iqr(sample(x, n, replace = TRUE),
                                       type = type))
  # Small n: a bootstrap resample can tie at the quartiles (q3=q1), making
  # lambda_iqr(.) = Inf. Drop these before averaging so one Inf doesn't
  # dominate the correction.
  boot_est <- boot_est[is.finite(boot_est)]
  2 * lambda_iqr(x, type = type) - mean(boot_est)
}
```

```{r estimators-demo}
x <- rexp(30, rate = 2)
c(ML = lambda_ml(x), IQR = lambda_iqr(x), bias_corrected = lambda_tilde(x))
```

## Monte Carlo comparison [20%]

**Design.** The structure follows the course's own Monte Carlo tutorial
[@frery2026mc]: for each situation $(\lambda, n) \in \Lambda \times N$,
$R$ independent replications are generated, each replication gives one
sample $\mathbf{X}$ of size $n$ and one value of each of the three
estimators, and the MSE for that situation is the sample mean of
$(\hat\lambda - \lambda)^2$ over the $R$ replications.

$R=5000$ is used for every situation here, straight from the Week 9
lecture, where the lecturer said 5000 replications per case is enough
for an assignment this size, and that you only need to control $R$ more
tightly in much bigger experiments. It also leaves a
decent margin on precision: the standard error of the Monte Carlo
estimate of MSE is about
$\sqrt{\widehat{\mathrm{Var}}[(\hat\lambda-\lambda)^2]/R}$, and even at
$n=5$, where the estimators are most variable, $R=5000$ keeps this small
next to the MSE values being compared.

```{r mc-experiment}
#| cache: true
mc_mse <- function(lambda, n, R, B = 200, type = 8) {
  est <- matrix(NA_real_, nrow = R, ncol = 3,
                dimnames = list(NULL, c("ML", "IQR", "bias_corrected")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, ] <- c(lambda_ml(x), lambda_iqr(x, type = type),
                  lambda_tilde(x, B = B, type = type))
  }
  colMeans((est - lambda)^2)
}

lambda_grid <- c(0.1, 1, 10)
n_grid <- c(5, 30, 100)
R <- 5000

results <- do.call(rbind, lapply(lambda_grid, function(lam) {
  do.call(rbind, lapply(n_grid, function(n) {
    mse <- mc_mse(lam, n, R = R)
    data.frame(lambda = lam, n = n, R = R, t(mse), check.names = FALSE)
  }))
}))
results
```

The same numbers are easier to compare as a plot, especially since
`bias_corrected`'s MSE spans several orders of magnitude across the
design -- both axes below are on a log scale.

```{r mc-plot}
#| fig-width: 8
#| fig-height: 3
est_names <- c("ML", "IQR", "bias_corrected")
results_long <- do.call(rbind, lapply(est_names, function(est) {
  data.frame(lambda = results$lambda, n = results$n,
             estimator = est, mse = results[[est]])
}))

ggplot(results_long, aes(n, mse, color = estimator)) +
  geom_line() + geom_point() +
  scale_x_log10(breaks = c(5, 30, 100)) +
  scale_y_log10() +
  facet_wrap(~ lambda, labeller = label_both) +
  labs(x = "n (log scale)", y = "MSE (log scale)", color = "Estimator")
```

**Discussion.** $\hat\lambda_{ML}$ has the lowest MSE in every one of the
nine cells. This matches theory. It is the maximum likelihood estimator,
so it uses the full sample through the likelihood, and it has the
smallest asymptotic variance among regular estimators. $\hat\lambda_{IQR}$
only uses two order statistics, so it should do worse. The MSE gap between
the two stays in a similar range across the design: $\hat\lambda_{IQR}$'s
MSE is roughly 1.6 to 2.5 times $\hat\lambda_{ML}$'s, at every
$(\lambda, n)$ combination, with no clear trend as $n$ or $\lambda$
change. This roughly matches the idea that the IQR estimator has a fixed
*asymptotic* efficiency loss from throwing away most of the sample, but
that result only has to hold as $n\to\infty$ -- some of the spread in the
ratio seen here could be genuine finite-sample behaviour rather than just
Monte Carlo noise from a design that is too small to pin the ratio down
exactly.

The bias-corrected estimator $\tilde\lambda$ is *worse* than the
uncorrected $\hat\lambda_{IQR}$ at every single $(\lambda, n)$ in the
design, not only at $n=5$. At $n=100$, the correction makes MSE about
15-18% bigger than $\hat\lambda_{IQR}$, across all three $\lambda$
values. At $n=30$, the gap is about 24-26%. At $n=5$, it is much worse:
three to four orders of magnitude bigger. The reason is a bias-variance
trade-off that goes the wrong way. $\hat\lambda_{IQR}$ is a nonlinear
function (a reciprocal) of the sample's interquartile spread. A bootstrap
resample can, just by chance, have a noticeably smaller or larger spread
than the original sample. After taking the reciprocal, and then doubling
it in $\tilde\lambda = 2\hat\lambda_{IQR} - \overline{\hat\lambda_{IQR}^*}$,
this small chance difference turns into a large swing in the correction
term. The variance this adds is bigger than the bias it removes, so MSE
gets worse overall, and more so as $n$ shrinks, because bootstrap
resamples are then more likely to have an unusually small spread. So the
first-order bootstrap correction does not help this estimator, at any of
the sample sizes tested here.

All three estimators' MSE decreases roughly with $1/n$, which matches the
usual parametric convergence rate, and it also increases roughly with
$\lambda^2$: at $n=5$, $\mathrm{MSE}(\hat\lambda_{ML})/\lambda^2$ is about
0.64, 0.57, and 0.62 at $\lambda=0.1,1,10$, and these are close enough
that the ratio looks roughly constant. That makes sense since $\hat\lambda$
estimates a rate (units of $1/\text{time}$), so squared error should
scale with $\lambda^2$ under a simple change of the time unit.

\newpage

# Exercise 2

```{r ausbeer}
autoplot(ausbeer) + labs(title = "Quarterly Australian beer production", y = "Megalitres")
```

## Frequency

TODO

## Trend

TODO

## Seasonality

TODO

## Variance stability

TODO

## Unusual observations

TODO

```{r ausbeer-season}
ggseasonplot(ausbeer)
```

## Seasonal plot

TODO

```{r ausbeer-subseries}
ggsubseriesplot(ausbeer)
```

## Seasonal subseries plot

TODO

## Detrend / transform?

TODO

# References
